Natural variation in teosinte at the domestication locus teosinte branched1 (tb1)

### 2014-02-16 manuscript.tex
edits throughout manuscript, comments for revamping discussion

\documentclass[12pt]{article}
\usepackage{geometry} 
\usepackage{hyperref}
\usepackage{color}
\usepackage[pdftex]{graphicx}  
%\usepackage{caption}
\usepackage{natbib}
\usepackage{authblk}
\renewcommand{\baselinestretch}{1.5}
\bibliographystyle{apalike}

\newcommand{\jri}[1]{\textcolor{blue}{ \emph{\scriptsize  #1}} }

@@

\author[1]{Laura Vann}
\author[1]{Thomas Kono}
\author[1]{Tanja Pyh\"aj\"arvi}
\author[1]{Matthew Hufford}
\author[1,2]{Jeffrey Ross-Ibarra\thanks{rossibarra@ucdavis.edu}}
\affil[1]{Department of Plant Sciences, University of California Davis}
\affil[2]{Center for Population Biology and Genome Center, University of California Davis}

@@

@@

Many morphological changes are associated with the domestication of maize, and understanding the genetic basis for these changes has been a focus of maize research for a number of years. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in branching differences between maize and teosinte \cite{Doebley et al 1990, Doebley and Stec 1991, Lukens and Doebley 1999}. Further studies have shown that the insertion of a 4.9 kb retrotransposon (Hopscotch) in the upstream control region of \emph{tb1} led to increased expression of this gene, causing the reduction in branching observed in domesticated maize \cite{Studer et al 2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in teosinte \cite{Studer et al 2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \cite{Studer et al 2011}. Furthermore, \cite{Studer and Doebley 2012} investigated the phenotypic effects of 9 teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte. In this study we aim to characterize the distribution of the Hopscotch insertion in \emph{mexicana}, \emph{parviglumis}, and landrace maize, and to examine the phenotypic effects of the insertion in \emph{parviglumis}. We use a combination of PCR genotyping for the Hopscotch element in our full panel and sequencing of two small regions upstream of \emph{tb1} in a subset of teosinte populations to explore patterns of genetic variation at this locus. Finally, we test for an association between the Hopscotch element and tillering phenotypes in a population of \emph{parviglumis}.

@@

In addition to genotyping, we chose a subset of \emph{parviglumis} individuals for sequencing. We chose twelve individuals from each of four populations from Jalisco state, Mexico (San Lorenzo, La Mesa, Ejutla A, and Ejutla B). For amplification and sequencing, we selected two regions approximately 600bp in size from within the 5' UTR of \emph{tb1} (sequenced region 1) and from 1,235 bp upstream of start of the Hopscotch and 66,169 bp upstream from the start of the \emph{tb1} ORF (sequenced region 2). The 5' UTR (containing sequenced region 1) has been shown to have elevated diversity in ssp. \emph{parviglumis} with respect to maize, while the the area upstream from the Hopscotch (sequenced region 2) has been shown to be critical in determining basal branching and ear architecture \cite{Wang et al 1999, Clark et al 2006}. We designed the following primers using PRIMER3 \cite{Rozen and Skaletsky 2000}: for the 5' UTR, 5' GGATAATGTGCACCAGGTGT 3' and 5' GCGTGCTAGAGACACYTGTTGCT 3'; for the 50 kb upstream region, 5' TGTCCTCGCCGCAACTC 3' and 5' TGTACGCCCGCCCCTCATCA 3' (Figure S1). We used Taq Polymerase (New England Biolabs) and the following thermal cycler conditions to amplify fragments: 94°C for 3 min, 30 cycles of 92°C for 40 s, annealing for 1 min, 72°C for 40 s, and a final 10 min extension at 72°C. Annealing temperatures for sequenced region 1 and sequenced region 2 were 59.7°C and 58.8°C, respectively. To clean excess primer and dNTPs we added two units of Exonuclease1 and 2.5 units of Antarctic Phosphatase to 8.0 µL of amplification product. This mix was placed on a thermal cycler with the following program: 37°C for 30 min, 80°C for 15 min, and a final cool-down step to 4°C.

@@

We examined discrepancies between observed and expected genotype frequencies by calculating Hardy-Weinberg Equilibrium (HWE), and to calculate differentiation between populations (FST) and subspecies (FCT) we used HierFstat \cite{Goudet 2005}. These analyses only included populations in which 8 or more individuals were sampled. To test the hypothesis that the Hopscotch insertion may be adaptive under certain environmental conditions, we looked for significant associations between the Hopscotch frequency and environmental variables using BayEnv \cite{Coop et al 2010}. BayEnv creates a covariance matrix of relatedness between populations, and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (Bayes Factor)\cite{Coop et al 2010}. We used genotyping and covariance data from Pyhäjärvi et al., (2013) \cite{Pyhajarvi et al 2013} for BayEnv, with the Hopscotch insertion coded as an additional SNP (Table S3). Environmental data were obtained from www.worldclim.org, the Harmonized World Soil Database and www.harvestchoice.org, and summarized by principle component analysis Pyhäjärvi et al. (2013) \cite{Pyhajarvi et al 2013}.

@@

For population genetic analyses of sequenced region 1 and sequenced region 2 we used the analysis package of Libsequence \cite{Thornton 2003} to calculate pairwise FST between populations, and to calculate standard diversity statistics (number of haplotypes; haplotype diversity; Watterson's estimator$\theta_{W}$; pairwise nucleotide diversity $\theta_{\pi}$; and Tajima's D). To produce a visual representation of differentiation between sequences and to examine patterns in sequence clustering by Hopscotch genotype we used Phylip (http://evolution.genetics.washington.edu/phylip.html) to create neighbor-joining trees with bootstrapping (100 repetitions) to examine the support of nodes in our trees. For creation of trees we also included homologous sequence data from teosinte inbred lines (TILs), some of which are known to be homozygous for the Hopscotch insertion (TIL03, TIL17, TIL09), as well as 59 lines of domesticated maize and landraces (data from Maize HapMapV2, \cite{Chia et al 2012}).

@@

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \cite{Falush et al 2003} and the same phased 55K SNP data from \cite{Pyhajarvi et al 2013} that we used for LD analysis, combined with the corresponding SNP data from a diverse panel of 282 maize lines \cite{Cook et al 2012}. SNPs were anchored in a modified version of the IBM genetic map ( \cite{Gerke et al 2013}, \url{http://arxiv.org/abs/1307.7313}). We created haplotype blocks using a custom Perl script that grouped all SNPs within a 5 kb window into a haplotype. We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

@@

We genotyped 247 accessions of maize landraces, and found the Hopscotch element fixed in all but 8 of these accessions (inbred lines X,Y, Z, and landraces in Chihuahua, Tapalpa, Tenancingo, and Altotonga) (Table S1, Table S2). For all analyses we only included individuals that clearly resolved for both PCRs, resulting in 837 individuals total. Within our \emph{parviglumis} and \emph{mexicana} samples we found the Hopscotch insertion segregating in 37 and 4 populations, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoacán in central-western Mexico in both subspecies (Figure 1). We examined Hardy-Weinberg equilibrium in a total of 14 populations (10 \emph{parviglumis} and 4 \emph{mexicana}) with more than 8 individuals sampled per population. Three populations (RIMPA0073, RIMPA0093, and RIMPA0158) show evidence of deviations from expected genotype frequencies under the assumptions of HWE (p=0.05).

Using our Hopscotch genotyping data, we calculated differentiation between populations (FST) and subspecies (FCT) for populations in which we sampled 8 or more alleles. FCT is 0 within our dataset, and we found similar levels of FST between populations within each subspecies (0.22) and between all populations (0.23), to those reported in genome-wide estimates from previous studies \cite{Pyhajarvi et al 2013} (Table 1). Although we found large variation in Hopscotch allele frequency among our populations, our BayEnv analysis did not indicate a correlation between the Hopscotch insertion and environmental variables (all Bayes Factors < 1; Table S3).

@@

To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the tb1 region, we sequenced two small (<1kb) regions upstream of the tb1 ORF in four populations. After alignment and singleton checking we recovered 40 and 48 segregating sites for the 50kb upstream region and the 5' UTR region, respectively. For region 1, Ejutla A, Ejutla B, and La Mesa have comparable values for haplotype diversity,$\theta_{W}$, $\theta_{\pi}$, and Tajima's D, while San Lorenzo has lower values for haplotype diversity,$\theta_{W}$, $\theta_{\pi}$, and a more positive value for Tajima's D (Table 2). For region 2, haplotype diversity,$\theta_{W}$, and $\theta_{\pi}$, are similar for Ejutla A and Ejutla B, while La Mesa and San Lorenzo have slightly lower values for these statistics (Table 2). Tajima's D is positive in all populations except San Lorenzo, indicating an excess of low frequency variants in this latter population (Table 2). Pairwise values of FST within population pairs Ejutla A/Ejutla B and San Lorenzo/La Mesa are 0 for both sequenced regions as well as for the Hopscotch, while they are high for other population pairs (Table 1). Neighbor joining trees of our sequence data and data from the teosinte inbred lines (TILS; data from Maize HapMapV2, \cite{Chia et al 2012}) do not reveal any clear clustering pattern with respect to population or Hopscotch genotype (Figure S3). Furthermore, individuals within our sample that have the Hopscotch insertion do not group with the teosinte inbred lines or the lines of domesticated maize that have the Hopscotch insertion.

@@

The lack of clustering of Hopscotch genotypes in our NJ tree as well as the lack of LD around \emph{tb1} does not support the hypothesis that the Hopscotch insertion in these populations of \emph{parviglumis} is the result of recent introgression. However, to further explore this hypothesis we performed a STRUCTURE analysis using Illumina MaizeSNP50 data from four of our \emph{parviglumis} populations (EjuA, EjuB, MSA, and SLO) and the maize 282 diversity panel \cite{Cook et al 2012, Phyhajarvi et al 2013}. The linkage model implemented in STRUCTURE can be used to identify ancestry of blocks of linked variants, which would arise as a result of recent admixture between populations. If the Hopscotch insertion is present in populations of \emph{parviglumis} as a result of recent admixture with domesticated maize, we would expect the insertion and linked variants in surrounding sites to be assigned to the "maize" cluster in our STRUCTURE runs, not the "teosinte" cluster. In all runs, assignment to maize in the \emph{tb1} region across all four \emph{parviglumis} populations is lower (average 0.017) than the chromosome wide average (0.20). (Figure 3).

@@

Adaptation occurs either due to selection on standing variation or on de novo mutations. Adaptation as a result of selection on standing variation has been well-described in a number of systems, for example, selection on lactase tolerance in humans \cite{Plantinga et al 2012, Tishkoff et al 2007}; variation at the \jri{italics?} Eda locus in three-spined stickleback \cite{Kitano et al 2008, Colosimo et al 2005}; and pupal diapause in the Apple Maggot fly \cite{Feder et al 2003}. Although the role of standing variation with respect to adaptation has been described in many systems, its importance to domestication is not as well studied.

In maize, alleles at important domestication loci (RAMOSA1, \cite{Sigmon and Vollbrecht 2010}; barren stalk1, \cite{Gallavotti et al 2004}; and \jri{do gene names need to be italicized?} grassy tillers1, \cite{Whipple et al 2011} have been shown to have been selected from standing variation, suggesting that diversity already present in teosinte may have played an important role in the domestication of maize. The \emph{teosinte branched1} gene has long been a central focus of research concerning maize domestication, and, while previous studies have suggested that differences in plant architecture between domesticated maize and teosinte are a result of selection on standing variation, little is known about variation at this locus in teosinte \cite{Clark et al 2006, Studer et al 2011}. Studer et al. (2011) genotyped 90 accessions of teosinte (inbred and non-inbred), and provided the first evidence that the Hopscotch insertion is segregating in teosinte \cite{Studer et al 2011}.

Given that the Hopscotch insertion has been estimated to predate the domestication of maize, it is not surprising that it is still segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the Hopscotch in teosinte. While our findings are consistent with a previous study by Studer et al. (2011) in that we identified the Hopscotch allele segregating in teosinte, we find it at higher frequency than previously suggested \cite{Studer et al 2011}. Crop to wild introgression has previously been observed at domestication loci (\cite{Ellstrand et al 1999, Zhang et al 2009, Thurber et al 2010, Baack et al 2008, Hubner et al 2012, Wilkes et al 1977, Van Heerwaarden et al 2011, Barrett 1983}), but our results are more consistent with Hufford et al. (2013) who found resistance to introgression from maize into teosinte \cite{Hufford et al 2013}. Furthermore, \jri{all citations, even citations like these need to reference the bibtex file. check out the difference betweetn cite, citet, and citep as commands. see \url{http://merkel.zoneo.net/Latex/natbib.php} for details} Hufford et al. (2013) showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize).

We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte. We do not have sympatric maize samples from many of our populations, however, so we cannot rule out the possibility that Hopscotch alleles in these teosinte populations are segregating in sympatric maize. On the other hand, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1. The similarity in LD in the \emph{tb1} region in populations with and without the Hopscotch provides further support for an explanation other than introgression. Additionally, assignment to maize in this region in our STRUCTURE analysis is in fact lower than the average across chromosome 1 (Figure 3, Table 4). \jri{figure references need to be as in the example using the ref command}

Although recent introgression seems unlikely, we cannot rule out ancient introgression as an explanation for the presence of the Hopscotch in these populations. In the ancient introgression scenario, recombination would have broken up linked variants surrounding the Hopscotch and we would find LD patterns similar to the ones we report here. However, it seems unlikely that there would have been ancient introgression between these populations of teosinte and their sympatric maize populations and not present-day introgression. Studies have demonstrated the presence of ongoing gene flow between domesticated maize and both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis} in a number of sympatric populations \cite{Hufford et al 2013, Ellstrand et al 2007, van Heerwaarden et al 2011}. \jri{check these refs. Hufford 2013 is mexicana, as is i think Ellstrand. do we show introgression with parv in van Heerwaarden?. do is how it in my 2009 paper?}

The lower levels of sequence diversity in the 5' UTR region and the more positive values of Tajima's D in the San Lorenzo population are consistent with patterns expected under neutrality. \jri{lower diversity is consistent with neutrality?} Though, deviations from expected genotype frequencies under Hardy-Weinberg equilibrium at the Hopscotch insertion in the San Lorenzo population and additional populations may be indicative of demographic changes that have affected genome-wide patterns of variation, or may be the result of selective pressures acting on or near the \emph{tb1} locus. \jri{shouldn't Tanja's data be able to argue one way or the other here? she looked at eveidence of inbreeding. what did she find in san lorenzo?} Similarly, if the Hopscotch is segregating as neutral variation in teosinte, the probability that it would have simultaneously drifted to high frequency in multiple populations is quite low; however, we find no evidence of selection or introgression at or near the \emph{tb1} locus in our sample set.

The phenotypic effects of the Hopscotch insertion in domesticated maize have been well documented \cite{Clark et al 2006, Studer et al 2011}, and Weber et al. (2007) have described its effects in partially inbred lines of teosinte \cite{Weber et al 2007}. Our study is the first to explicitly examine the phenotypic effects of the Hopscotch insertion in individuals sampled from a natural population of teosinte. However, we found no significant effect of the Hopscotch on tillering index or tiller number in our phenotyping experiments, and the effect of the Hopscotch insertion in teosinte is discordant with that of maize. The lack of correlation between Hopscotch genotype and tillering index or tiller number is surprising given its effects in maize, as well as our own preliminary data collected on a grow-out of teosinte inbred lines. \jri{Huff, think we should include the TIL data just to show we can see effects in our greenhouse?}

>>>>

\documentclass[12pt]{article}
\usepackage{geometry} 
\usepackage{hyperref}
\usepackage{color}
\usepackage[pdftex]{graphicx}  
%\usepackage{caption}
\usepackage{natbib}
\usepackage{authblk}
\usepackage{subscript}
\renewcommand{\baselinestretch}{1.5}
\bibliographystyle{apalike}

\newcommand{\jri}[1]{\textcolor{blue}{ \emph{\scriptsize  #1}} }
\newcommand{\mbh}[1]{\textcolor{red}{ \emph{\scriptsize  #1}} }

@@

\author[1]{Laura Vann}
\author[1]{Thomas Kono}
\author[1]{Tanja Pyh\"aj\"arvi}
\author[1]{Matthew B. Hufford}
\author[1,2]{Jeffrey Ross-Ibarra\thanks{rossibarra@ucdavis.edu}}
\affil[1]{Department of Plant Sciences, University of California Davis}
\affil[2]{Center for Population Biology and Genome Center, University of California Davis}

@@

\mbh{my new address is: Department of Ecology, Evolution, and Organismal Biology, Iowa State University, Ames, IA, 50011}

@@

Many morphological changes are associated with the domestication of maize, and understanding the genetic basis for these changes has been a focus of maize research for a number of years. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in branching differences between maize and teosinte \cite{Doebley et al 1990, Doebley and Stec 1991, Lukens and Doebley 1999}. Further studies have shown that the insertion of a 4.9 kb retrotransposon (Hopscotch) in the upstream control region of \emph{tb1} led to increased expression of this gene, causing the reduction in branching observed in domesticated maize \cite{Studer et al 2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in teosinte \cite{Studer et al 2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \cite{Studer et al 2011}. Furthermore, \cite{Studer and Doebley 2012} investigated the phenotypic effects of 9 teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte. \mbh{A sentence or two is needed here explaining why tillering may be an ecologically relevant trait to a wild plant} In this study we aim to characterize the distribution of the Hopscotch insertion in \emph{mexicana}, \emph{parviglumis}, and landrace maize, and to examine the phenotypic effects of the insertion in \emph{parviglumis}. We use a combination of PCR genotyping for the Hopscotch element in our full panel and sequencing of two small regions upstream of \emph{tb1} in a subset of teosinte populations to explore patterns of genetic variation at this locus. Finally, we test for an association between the Hopscotch element and tillering phenotypes in a population of \emph{parviglumis}.

@@

In addition to genotyping, we chose a subset of \emph{parviglumis} individuals for sequencing. We chose twelve individuals from each of four populations from Jalisco state, Mexico (San Lorenzo, La Mesa, Ejutla A, and Ejutla B). For amplification and sequencing, we selected two regions approximately 600bp in size from within the 5' UTR of \emph{tb1} (sequenced region 1) and from 1,235 bp upstream of the start of the Hopscotch and 66,169 bp upstream from the start of the \emph{tb1} ORF (sequenced region 2). The 5' UTR (containing sequenced region 1) has been shown to have elevated diversity in ssp. \emph{parviglumis} with respect to maize, while the the area upstream from the Hopscotch (sequenced region 2) has been shown to be critical in determining basal branching and ear architecture \cite{Wang et al 1999, Clark et al 2006}. We designed the following primers using PRIMER3 \cite{Rozen and Skaletsky 2000}: for the 5' UTR, 5' GGATAATGTGCACCAGGTGT 3' and 5' GCGTGCTAGAGACACYTGTTGCT 3'; for the 50 kb upstream region, 5' TGTCCTCGCCGCAACTC 3' and 5' TGTACGCCCGCCCCTCATCA 3' (Figure S1). We used Taq Polymerase (New England Biolabs) and the following thermal cycler conditions to amplify fragments: 94°C for 3 min, 30 cycles of 92°C for 40 s, annealing for 1 min, 72°C for 40 s, and a final 10 min extension at 72°C. Annealing temperatures for sequenced region 1 and sequenced region 2 were 59.7°C and 58.8°C, respectively. To clean excess primer and dNTPs we added two units of Exonuclease1 and 2.5 units of Antarctic Phosphatase to 8.0 µL of amplification product. This mix was placed on a thermal cycler with the following program: 37°C for 30 min, 80°C for 15 min, and a final cool-down step to 4°C.

@@

We examined discrepancies between observed and expected genotype frequencies by calculating Hardy-Weinberg Equilibrium (HWE), and to calculate differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) we used HierFstat \cite{Goudet 2005}. These analyses only included populations in which 8 or more individuals were sampled. To test the hypothesis that the Hopscotch insertion may be adaptive under certain environmental conditions, we looked for significant associations between the Hopscotch frequency and environmental variables using BayEnv \cite{Coop et al 2010}. BayEnv creates a covariance matrix of relatedness between populations, and then tests a null model that allele frequencies in populations are determined by the covariance matrix of relatedness alone against the alternative model that allele frequencies are determined by a combination of the covariance matrix and an environmental variable, producing a posterior probability (Bayes Factor)\cite{Coop et al 2010}. We used genotyping and covariance data from Pyhäjärvi et al., (2013) \cite{Pyhajarvi et al 2013} for BayEnv, with the Hopscotch insertion coded as an additional SNP (Table S3). Environmental data were obtained from www.worldclim.org, the Harmonized World Soil Database and www.harvestchoice.org, and summarized by principle component analysis Pyhäjärvi et al. (2013) \cite{Pyhajarvi et al 2013}.

@@

For population genetic analyses of sequenced region 1 and sequenced region 2 we used the analysis package of Libsequence \cite{Thornton 2003} to calculate pairwise F\textsubscript{ST} between populations, and to calculate standard diversity statistics (number of haplotypes; haplotype diversity; Watterson's estimator $\theta_{W}$; pairwise nucleotide diversity $\theta_{\pi}$; and Tajima's D). To produce a visual representation of differentiation between sequences and to examine patterns in sequence clustering by Hopscotch genotype we used Phylip (http://evolution.genetics.washington.edu/phylip.html) to create neighbor-joining trees with bootstrapping (100 repetitions) to examine the support of nodes in our trees. For creation of trees we also included homologous sequence data from teosinte inbred lines (TILs), some of which are known to be homozygous for the Hopscotch insertion (TIL03, TIL17, TIL09), as well as 59 lines of domesticated maize and landraces (data from Maize HapMapV2, \cite{Chia et al 2012}).

@@

We examined evidence of introgression on chromosome 1 in these same four populations (EjuA, EjuB, MSA, SLO) using STRUCTURE \cite{Falush et al 2003} and the same phased 55K SNP data from \cite{Pyhajarvi et al 2013} that we used for LD analysis, combined with the corresponding SNP data from a diverse panel of 282 maize lines \cite{Cook et al 2012}. SNPs were anchored in a modified version of the IBM genetic map ( \cite{Gerke et al 2013}, \url{http://arxiv.org/abs/1307.7313}). We created haplotype blocks using a custom Perl script that grouped SNPs separated by less than 5kb into haplotypes. We ran STRUCTURE at K=2 under the linkage model, performing 3 replicates with an MCMC burn-in of 10,000 steps and 50,000 steps post burn-in.

@@

We genotyped 247 accessions of maize landraces, and found the Hopscotch element fixed in all but 8 of these accessions (inbred lines X,Y, Z, and landraces in Chihuahua, Tapalpa, Tenancingo, and Altotonga) (Table S1, Table S2). For all analyses we only included individuals that clearly resolved for both PCRs, resulting in 837 individuals total. Within our \emph{parviglumis} and \emph{mexicana} samples we found the Hopscotch insertion segregating in 37 and 4 populations, respectively, and at highest frequency in the states of Jalisco, Colima, and Michoac\'{a}n in central-western Mexico in both subspecies (Figure 1). We examined Hardy-Weinberg equilibrium in a total of 14 populations (10 \emph{parviglumis} and 4 \emph{mexicana}) with more than 8 individuals sampled per population. Three populations (RIMPA0073, RIMPA0093, and RIMPA0158) show evidence of deviations from expected genotype frequencies under the assumptions of HWE (p=0.05). \mbh{not sure what p=0.05 means here.  Should it be p $<$ 0.05 or $\alpha$ = 0.05?}

Using our Hopscotch genotyping data, we calculated differentiation between populations (F\textsubscript{ST}) and subspecies (F\textsubscript{CT}) for populations in which we sampled 8 or more alleles. F\textsubscript{CT} is 0 within our dataset, and we found similar levels of F\textsubscript{ST} between populations within each subspecies (0.22) and between all populations (0.23), to those reported in genome-wide estimates from previous studies \cite{Pyhajarvi et al 2013} (Table 1). Although we found large variation in Hopscotch allele frequency among our populations, our BayEnv analysis did not indicate a correlation between the Hopscotch insertion and environmental variables (all Bayes Factors $<$ 1; Table S3).

@@

To investigate patterns of sequence diversity and linkage disequilibrium (LD) in the tb1 region, we sequenced two small ($<$1kb) regions upstream of the tb1 ORF in four populations. After alignment and singleton checking we recovered 40 and 48 segregating sites for the 50kb upstream region and the 5' UTR region, respectively. For region 1, Ejutla A, Ejutla B, and La Mesa have comparable values for haplotype diversity, $\theta_{W}$, $\theta_{\pi}$, and Tajima's D, while San Lorenzo has lower values for haplotype diversity, $\theta_{W}$, $\theta_{\pi}$, and a more positive value for Tajima's D (Table 2). For region 2, haplotype diversity, $\theta_{W}$, and $\theta_{\pi}$, are similar for Ejutla A and Ejutla B, while La Mesa and San Lorenzo have slightly lower values for these statistics (Table 2). Tajima's D is positive in all populations except San Lorenzo, indicating an excess of low frequency variants in this latter population (Table 2). \mbh{this text hasn't changed since the last time I read the manuscript and previously did not agree with the results presented in Table 2.  Was Table 2 incorrect?} Pairwise values of F\textsubscript{ST} within population pairs Ejutla A/Ejutla B and San Lorenzo/La Mesa are 0 for both sequenced regions as well as for the Hopscotch, while they are high for other population pairs (Table 1). Neighbor joining trees of our sequence data and data from the teosinte inbred lines (TILS; data from Maize HapMapV2, \cite{Chia et al 2012}) do not reveal any clear clustering pattern with respect to population or Hopscotch genotype (Figure S3). Furthermore, individuals within our sample that have the Hopscotch insertion do not group with the teosinte inbred lines or the lines of domesticated maize that have the Hopscotch insertion.

@@

The lack of clustering of Hopscotch genotypes in our NJ tree as well as the lack of LD around \emph{tb1} does not support the hypothesis that the Hopscotch insertion in these populations of \emph{parviglumis} is the result of recent introgression. However, to further explore this hypothesis we performed a STRUCTURE analysis using Illumina MaizeSNP50 data from four of our \emph{parviglumis} populations (EjuA, EjuB, MSA, and SLO) and the maize 282 diversity panel \cite{Cook et al 2012, Phyhajarvi et al 2013}. The linkage model implemented in STRUCTURE can be used to identify ancestry of blocks of linked variants, which would arise as a result of recent admixture between populations. If the Hopscotch insertion is present in populations of \emph{parviglumis} as a result of recent admixture with domesticated maize, we would expect the insertion and linked variants in surrounding sites to be assigned to the "maize" cluster in our STRUCTURE runs, not the "teosinte" cluster. In all runs, assignment to maize in the \emph{tb1} region across all four \emph{parviglumis} populations is lower (average 0.017) than the chromosome wide average (0.20; Figure 3).

@@

Adaptation occurs either due to selection on standing variation or on \emph{de novo} mutations. Adaptation as a result of selection on standing variation has been well-described in a number of systems, for example, selection for lactose tolerance in humans \cite{Plantinga et al 2012, Tishkoff et al 2007}; variation at the \jri{italics?} Eda locus in three-spined stickleback \cite{Kitano et al 2008, Colosimo et al 2005}; and pupal diapause in the Apple Maggot fly \cite{Feder et al 2003}. Although the role of standing variation with respect to adaptation has been described in many systems, its importance to domestication is not as well studied.

In maize, alleles at important domestication loci (RAMOSA1, \cite{Sigmon and Vollbrecht 2010}; barren stalk1, \cite{Gallavotti et al 2004}; and \jri{do gene names need to be italicized?} \mbh{yup, they need to be italicized} grassy tillers1, \cite{Whipple et al 2011}) have been shown to have been selected from standing variation, suggesting that diversity already present in teosinte may have played an important role in the domestication of maize. The \emph{teosinte branched1} gene has long been a central focus of research concerning maize domestication, and, while previous studies have suggested that differences in plant architecture between domesticated maize and teosinte are a result of selection on standing variation, little is known about variation at this locus in teosinte \cite{Clark et al 2006, Studer et al 2011}. Studer et al. (2011) genotyped 90 accessions of teosinte (inbred and outbred), and provided the first evidence that the Hopscotch insertion is segregating in teosinte \cite{Studer et al 2011}.

Given that the Hopscotch insertion has been estimated to predate the domestication of maize, it is not surprising that it can be found segregating in populations of teosinte. However, in sampling numerous individuals from many teosinte populations our study provides greater insight into the distribution and prevalence of the Hopscotch in teosinte. While our findings are consistent with a previous study by Studer et al. (2011) in that we identified the Hopscotch allele segregating in teosinte, we find it at higher frequency than previously suggested \cite{Studer et al 2011}. Crop to wild introgression has previously been observed at domestication loci (\cite{Ellstrand et al 1999, Zhang et al 2009, Thurber et al 2010, Baack et al 2008, Hubner et al 2012, Wilkes et al 1977, Van Heerwaarden et al 2011, Barrett 1983}), but our results are more consistent with Hufford et al. (2013) who found resistance to introgression from maize into teosinte \cite{Hufford et al 2013}. Furthermore, \jri{all citations, even citations like these need to reference the bibtex file. check out the difference betweetn cite, citet, and citep as commands. see \url{http://merkel.zoneo.net/Latex/natbib.php} for details} Hufford et al. (2013) showed that domestication loci, such as \emph{tb1}, are particularly resistant to introgression in both directions of gene flow (i.e., maize to teosinte and teosinte to maize).

We find no evidence of recent introgression in our analyses. Clustering patterns in our NJ trees do not reflect a pattern expected if maize alleles at the \emph{tb1} locus had introgressed into populations of teosinte. We do not have sympatric maize samples from many of our populations, however, so we cannot rule out the possibility that Hopscotch alleles in these teosinte populations are segregating in sympatric maize. Moreover, analysis of linkage in the \emph{tb1} region does not reveal patterns of high LD relative to the rest of chromosome 1. The similarity in LD in the \emph{tb1} region in populations with and without the Hopscotch provides further support for an explanation other than introgression. Additionally, assignment to maize in this region in our STRUCTURE analysis is in fact lower than the average across chromosome 1 (Figure 3, Table 4). \jri{figure references need to be as in the example using the ref command}

Although recent introgression seems unlikely, we cannot rule out ancient introgression as an explanation for the presence of the Hopscotch in these populations. In the ancient introgression scenario, recombination would have broken up linked variants surrounding the Hopscotch and we would find LD patterns similar to the ones we report here. However, it seems unlikely that there would have been ancient introgression between these populations of teosinte and their sympatric maize populations and not present-day introgression. Studies have demonstrated the presence of ongoing gene flow between domesticated maize and both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis} in a number of sympatric populations \cite{Hufford et al 2013, Ellstrand et al 2007, van Heerwaarden et al 2011}. \jri{check these refs. Hufford 2013 is mexicana, as is i think Ellstrand. do we show introgression with parv in van Heerwaarden?. how does it look in my 2009 paper?} \mbh{there is a low level of introgression between maize and \emph{parviglumis} based on the results from van Heerwaarden, much less than between \emph{mexicana} and maize}

The lower levels of sequence diversity in the 5' UTR region and the more positive values of Tajima's D in the San Lorenzo population are consistent with patterns expected under neutrality. \jri{lower diversity is consistent with neutrality?} \mbh{this is probably coming from my last round of edits; LV, my comment was that the Taj D results in Table 2 were more consistent with neutrality in San Lorenzo rather than weak selection as it was previously written in the manuscript.  Perhaps the diversity is low because this is just a low-diversity population.  My SSR data showed evidence of a pop. bottleneck in SLO...think that's what we're seeing at tb1 too} Though, deviations from expected genotype frequencies under Hardy-Weinberg equilibrium at the Hopscotch insertion in the San Lorenzo population and additional populations may be indicative of demographic changes that have affected genome-wide patterns of variation, or may be the result of selective pressures acting on or near the \emph{tb1} locus. \jri{shouldn't Tanja's data be able to argue one way or the other here? she looked at eveidence of inbreeding. what did she find in san lorenzo?} \mbh{she found evidence of inbreeding in SLO as did I; you can cite both her paper and my dissertation for this if you'd like} Similarly, if the Hopscotch is segregating as neutral variation in teosinte, the probability that it would have simultaneously drifted to high frequency in multiple populations is quite low; however, we find no evidence of selection or introgression at or near the \emph{tb1} locus in our sample set. \mbh{many populations in the Jalisco cluster of Fukunaga seem to have high frequency of the hopscotch allele.  Why?  Only ancient introgression in Jalisco and not in the Balsas? If hopscotch came from standing variation, is this more consistent with a Jalisco domestication?  I know, heresy!!  LV, can you look at the \emph{parviglumis} populations with high hopscotch and see if they are all in the Jalisco cluster of Fukunaga?  Might be a nice observation we could make here...could even go so far as saying the results from Matsuoka deserve another look.}
